---
title: "Data Challenge #1"
author: "Mario Saraiva -mhs2195"
output: 
  html_document: 
    code_folding: hide
    toc: TRUE
    toc_float: TRUE
---

```{r}
##About this Project
# File-Name:      Data_Challenge_1_MarioSaraiva.Rmd
# Version:        R 3.4.1
# Date:           Feb 14, 2018
# Author:         MS
# Purpose:        Verify the lethality index with reported figures reported by 
# the NY Times in 2016 that said: " `86.1% of dead civilians` who presumably 
# participated in confrontations with federal armed forces were killed in events 
# of "perfect lethality" where there were only dead and no wounded. [...] Mexico 
# has the terrible situation of having `lethality indices of 2.6`. The lethality 
# index of the Federal Police is `2.6 dead for every wounded`, the Navy's reaches
# 17.3 dead for every wounded, and the Army's is 9.1 dead for every wounded."
# (Ahmed & Schmitt, 2016)
# Input Files:    1. AllViolenceData_171220.csv
# Output Files:   Data_Challenge_1_MarioSaraiva.html
# Data Output:    None
# Previous files: None
# Dependencies:   None
# Required by:    None
# Status:         IN PROGRESS
# Machine:        Mac laptop
```

Please address the following:


1. ~~Can you replicate the 86.1% number? the overall lethality ratio? the ratios for the Federal Police, Navy and Army?~~
    
    a. Provide a visualization that presents this information neatly. 
    
    ~~b. Please show the exact computations you used to calculate them.~~
  

 ~~ 2. Is this the right metric to look at? Why or why not?~~
    
  ~~3. What is the "lethality index" showing explicitly? What is it not showing? What is the definition assuming?~~
    
  ~~4. With the same available data, can you think of an alternative way to capture the same construct? Is it "better"?~~
    
  ~~5. What additional information would you need to better understand the data?~~
    
  ~~6. What additional information could help you better capture the construct behind the "lethality index"~~

-----  

```{r Libraries, warning=FALSE, message=FALSE, echo=FALSE}
#Clear all previous output
rm(list = ls())

#Libraries used in this project
library(readr)
library(ggplot2)
library(dplyr)
library(mice)
library(VIM)
```

```{r Load Data, warning=FALSE, message=FALSE, echo=FALSE}
#Load the data
dta <- read_csv("~/Documents/GitHub/First-Repo/Data Challenge 1/data/processed/AllViolenceData_171220.csv")
#dim(dta)
#str(dta)

##Convert the date into 3 collumns to ease visualizations. 
datetxt <- dta$date
datetxt <- as.Date(datetxt)
dates_df <- data.frame(date = datetxt,
                 year = as.numeric(format(datetxt, format = "%Y")),
                 month = as.numeric(format(datetxt, format = "%m")),
                 day = as.numeric(format(datetxt, format = "%d")))

dta <- right_join(dta,dates_df)
```

##Lethality Rate in Mexico 
  
<p> Stats given in the NY Times article: </p>
1. `86.1% of dead civilians`  in confrontations with __federal armed forces__.
2. `Mexico's lethality indices of 2.6`
3. `2.6 dead for every wounded` ratio for the federal POLICE.
4. `17.3 dead for every wounded` ratio for the NAVY.
5. `9.1 dead for every wounded` ratio for the ARMY.

The NY Times article did not discuss how these numbers were calculated. I explored different ways they could have calculated to arrive at these numbers. To no one's suprise, my results were different than the NY times results.

```{r Aggregate for Armed Forces}
#The proportion for the aggregate civilians lethality rate was the following

dta_gov_nongov <- dta %>% 
  mutate(armforces_death = navy_dead + military_dead + federal_police_dead + state_police_dead + municipal_police_dead + ministerial_police_dead,
         armforces_wounded = navy_wounded + military_wounded + federal_police_wounded + state_police_wounded + municipal_police_wounded + ministerial_police_wounded,  
         nonfed_wounded = civilian_wounded + organized_crime_wounded + afi_wounded + public_prosecutor_wounded,
         nonfed_dead = civilian_dead + organized_crime_dead + afi_dead + public_prosecutor_dead)
```

```{r Armed Forces Rate}
#Presence of Armed forces and the impact on civilians. 

##The following will count the total number of dead non-gov subjects given the presence of Armed Forces (A.F.)

indexes <- dta_gov_nongov %>%
  filter(armforces_wounded != 0 | armforces_death != 0 ) %>% #to make sure the armed forces were involved.
  summarize(count = sum(nonfed_dead))
colnames(indexes) <- "All.Non.AF_dead|A.F."
#The following will count the total number of wounded non-gov subjects given the presence of Armed Forces

indexes$All.NonAF.wound <- as.numeric(dta_gov_nongov %>%
  filter(armforces_wounded != 0 | armforces_death != 0 ) %>%
  summarize(count = sum(nonfed_wounded)))
#colnames(indexes$All.NonAF.wound) <- "All.Non.AF_wounded|A.F."

indexes$All.index <- as.numeric(indexes$`All.Non.AF_dead|A.F.` / indexes$All.NonAF.wound)
#colnames(indexes$All.index) <- "Armed Forces Lethality index "

count_all_total <- as.numeric(dta_gov_nongov %>%
   summarize(count = sum(armforces_death)))

indexes$All.Prop <- as.numeric(indexes$`All.Non.AF_dead|A.F.`/count_all_total)
#colnames(indexes$All.Prop) <- "non AF dead / AF dead"

```

```{r Perfect Lethality cases}

#Perfect Lethality cases
indexes$Perfect.L.dead <- as.numeric(dta_gov_nongov %>%
     filter(perfect_lethality != 0 & civilian_dead != 0) %>%
   summarize(count = sum(nonfed_dead)))

indexes$Perfect.L..Index <- as.numeric(indexes$Perfect.L.dead / (indexes$`All.Non.AF_dead|A.F.` + indexes$All.NonAF.wound))

```

```{r Federal Police, message=FALSE, warning=FALSE}
#Lethality ratios.

##Similarly, this section will calculate the 'lethality index' (num. of civilians dead / num. of civilians wounded) given the participation of each armed force in the violent. I also calculate the proportion of civilians that died given the number of armed force officers that die, (Total Num. deaths non-armed forces) / (Total Num. deaths armed forces). 

###This index has a clear shortcoming for it does not consider any armed forces officers that were dead or wounded during the violent event. This lethality index cannot be used to assess brutality of armed forces. 

#Federal Police
##1 Aggregate for non A.F.
indexes$FP.nonAF.dead <- as.numeric(dta_gov_nongov %>%
   filter(federal_police_dead != 0 | federal_police_wounded != 0) %>%
   summarize(count = sum(nonfed_dead)))
##2
indexes$FP.nonAF.wounded <- as.numeric(dta_gov_nongov %>%
   filter(federal_police_dead != 0 | federal_police_wounded != 0) %>%
   summarize(count = sum(nonfed_wounded)))
##3
indexes$FP.nonAF.Index <- as.numeric(indexes$FP.nonAF.dead / (indexes$FP.nonAF.dead + indexes$FP.nonAF.wounded))
#colnames(indexes$FP.Index) <- "Lethality index | Federal Police "

##4 - Aggregates for Federal Police officers
indexes$FP.only.dead <- as.numeric(dta_gov_nongov %>%
  filter(federal_police_dead != 0 | federal_police_wounded != 0) %>%
   summarize(count = sum(federal_police_dead)))
##5
indexes$FP.only.wounded <- as.numeric(dta_gov_nongov %>%
  filter(federal_police_dead != 0 | federal_police_wounded != 0) %>%
   summarize(count = sum(federal_police_wounded)) )
##6
indexes$FP.only.Index <- as.numeric(indexes$FP.only.dead / (indexes$`All.Non.AF_dead|A.F.` + indexes$All.NonAF.wound) )
#colnames(indexes$FP.only.Index) <- "FP only dead/wounded "

##7 Ratio of total non AF dead / total FP dead
indexes$FP.prop.of.dead <- as.numeric(indexes$`All.Non.AF_dead|A.F.` / indexes$FP.only.dead)


display_index <- t(format(round(indexes, 2), nsmall = 2))
display_index <- as.data.frame(display_index)
#colnames(display_index)[1] <- "Results"

```


```{r Navy and Army, message=FALSE, warning=FALSE}
#Navy
##1 Aggregate for non A.F. | Navy presence
indexes$NV.nonAF.dead <- as.numeric(dta_gov_nongov %>%
   filter(navy_dead != 0 | navy_wounded != 0) %>%
   summarize(count = sum(nonfed_dead)))
##2
indexes$NV.nonAF.wounded <- as.numeric(dta_gov_nongov %>%
    filter(navy_dead != 0 | navy_wounded != 0) %>%                                     
   summarize(count = sum(nonfed_wounded)))
##3
indexes$NV.nonAF.Index <- as.numeric(indexes$NV.nonAF.dead / (indexes$NV.nonAF.dead + indexes$NV.nonAF.wounded))
#colnames(indexes$FP.Index) <- "Lethality index | Navy "

##4 - Aggregates for Navy officers
indexes$NV.only.dead <- as.numeric(dta_gov_nongov %>%
  filter(navy_dead != 0 | navy_wounded != 0) %>%   
   summarize(count = sum(navy_dead)))
##5
indexes$NV.only.wounded <- as.numeric(dta_gov_nongov %>%
  filter(navy_dead != 0 | navy_wounded != 0) %>%   
   summarize(count = sum(navy_wounded)) )
##6
indexes$NV.only.Index <- as.numeric(indexes$NV.only.dead / (indexes$`All.Non.AF_dead|A.F.` + indexes$All.NonAF.wound))
#colnames(indexes$FP.only.Index) <- "FP only dead/wounded "

##7 Ratio of total non AF dead / total NV dead
indexes$NV.prop.of.dead <- as.numeric(indexes$`All.Non.AF_dead|A.F.` / indexes$NV.only.dead)

#Army
##No data on number of army officers wounded or killed.

#display_index <- t(format(round(indexes, 2), nsmall = 2))
#colnames(display_index)[1] <- "Results"

```

According to the NY times `86.1% of dead civilians`  in confrontations with __federal police__ . So I tried replicating this result through the follwoing, where non armed forces can be interpret as civilians:

$$\frac{ Total \space Num. \space of \space non \space Armed\space Forces \space subjects \space dead \space | \space Armed \space Forces \space presence \space}{Total \space Num. \space of \space non \space Armed\space Forces \space subjects \space (wounded+dead)\space | \space Armed \space Forces \space presence}$$  $$= \frac{1259}{1259 + 368} * 100 = 76.90\%$$

See the next section for an explanation on the computations used. According to my calculations, given the data, the ratio of $\frac{dead}{wounded}$ for all violent events is `r format(round(indexes$All.Prop, 2), nsmall =2)`. For every wounded during violent events there was at least 1 dead.

86.1% of dead civilians in confrontations with federal armed forces.
Mexico's lethality indices of 2.6
2.6 dead for every wounded ratio for the federal POLICE.
17.3 dead for every wounded ratio for the NAVY.
9.1 dead for every wounded ratio for the ARMY

```{r}
nytimes <- 0.86
nytimes <- as.data.frame(nytimes)
nytimes$All.index <- 2.6
nytimes$FP.prop.dead <- 2.6
nytimes$NV.prop.dead <- 17.3
nytimes$Army.prop.dead <- 9.1

compareA <- indexes$All.index
compareA <- as.data.frame(compareA)
compareA$FP.prop.dead <- indexes$FP.prop.of.dead
compareA$NV.prop.dead <- indexes$NV.prop.of.dead
compareA$Army <- "NA"

#indexes$FP.nonAF.Index <- format(round(indexes$FP.nonAF.Index, 2), nsmall = 2)

compare_indexes <- rbind(nytimes, c(indexes$FP.nonAF.Index, indexes$All.index, indexes$FP.prop.of.dead, indexes$NV.prop.of.dead, "NA"))
row.names(compare_indexes)[1] <- "New York times"
row.names(compare_indexes)[2] <- "Results"
compare_indexes$Source <- c("New York times", "Results")

compare_indexes$nytimes <- as.numeric(compare_indexes$nytimes)

ggplot(compare_indexes) +
  geom_col(aes(Source, nytimes, fill = Source)) +
  ylim(0, 1) + ggtitle("Lethality index for the Federal Police") +
  labs(x = "Source", y = "Lethality Index") +
  geom_text(aes(Source, nytimes, label = nytimes), position = position_stack(vjust = 0.5))
  
```

##Discussion of Alertnative Calculations

<p> Using the data avaialble, I defined the lethality index can be defined as the ratio of the total number of civillians^1^ dead given the presence of the armed forces^2^ over the total number of civilians wounded given the presence of the armed forces.
$\frac{ Num. \space of \space dead \space}{Num.\space of \space wounded\space}$ 
per confrontation or event. But this is confusing. For example
the overall lethality of government armed forces should be:

$$\frac{ Total \space Num. \space of \space non \space Armed\space Forces \space subjects \space dead \space | \space Armed \space Forces \space presence \space}{Total \space Num. \space of \space non \space Armed\space Forces \space subjects \space wounded\space | \space Armed \space Forces \space presence}$$. 

So given the presence of at least one government agency what is the proportion of dead and wounded people during the violent event that are not considered part of the armed forces? Confontations vary a lot and are complex. For instance, battling in the sea is very different than on ground, so many wounded in the sea does not have the same implications as wounded in land confrontations. Therefore, we should not over-strecth this ratio as an indicator of brutality of federal armed forces. This ratio also does not address brutality of armed forces given no confrontation nor does it take into account the guns used; 100 wounded using a heavy automatic weaponry is not the same as 100 wounded with a knife or physical agressions. </p>

Alternative calculations measures what happened given the presence of a certain group (e.g. Federal Police) engaging in a confrontation. In other words, it measures the ratio of Federal Police officers engaged in a confrontation that died in comparison to other officers that were wounded. Then we can compare this ratio for Federal Police and Civilians. Although, that still does not shows us much.


###Computations

Through seven computations I investigated the lethality index for Mexico, 
the Federal Police, Navy, and Army, as well as, the dead of civilians per death of an  agency's officer. I will use the calculations performed for the Navy as an example of these seven calculations. All other calculations follow a similar pattern.

I developed alternatives attempting to overcomer the limitations imposed by the data. These can be seen in my calculations.

__Calculations:__

1. First, I calculated the aggregate the number of non Armed Forces dead given 
the presence of the Navy in the violent event. I did not feel comfortable using
the binary variables provided indicating the presence of an armed force or not.
To produceconservative estimates I used only observations that had either one 
dead from the navy or one wounded. 
  
2. Similarly, I calculated the aggregate number of non Armed Forces wounded.
  
3. Then I calculated the proportion of total Non Armed forces dead over total 
Non Armed forces wounded. This is a lethality index capturing conflicts with the
presence of the Navy.

4. Forth, I calculated the aggregate number of Navy officers dead in violent 
events.

5. Like in step 4,  I calculated the aggregate number of Navy officers wounded 
in violent events.

6. At this stage, I calculated the number of non armed forces dead given the 
presence of the Navy in the violent event over the total amount of Non Armed 
Forces people in the event.

7. Lastly, I calculated the ratio of Total number of Non Armed Forces dead over the 
total number of navy officers dead. This gives us the ratio of dead civilians per
death in the navy.


##Results

####Data visualizations

```{r, message=FALSE, warning=FALSE}

# ggplot(df_rates) + 
#   geom_col(aes(x = V2, y = V1, fill = V2)) + 
#   labs(x = "Type of people", y = "Proportion") + 
#   ggtitle("Proportion of Lethality (Num. deaths | wounded = 0) / (Total deaths) by people type")
#   # ggsave(file = "plot1.png")
```

#### Summary Table

Variables         | Results       | New York Times
----------------- | ------------- | -------------
Perfect Leth Index| 9.15%         | 
Number of Non Armed Forces dead       | 8608          |
Number of Non Armed Forces wounded    | 3350          |
Lethality index for Mexico| 2.57          | 2.60
Dead-to-Wounded Ratio| 1.25          | 30
Number of non AF dead given Federal Police    | 1259          |
Number of non AF wounded given Federal Police | 378           |
Lethality Index for Federal Police | 3.33          |86%
Number of Federal Police Officers dead  | 872           |
Number of Federal Police Officers wounded|2902          |
Federal Police officers dead-to-wound ratio | 0.30%         |
Non Armed Forces Dead-to-Wound Ratio given Federal Police | 9.87          | 2.6
Number of non Armed Forces dead given Navy    | 800           |
Number of non Armed Forces wounded given Navy | 69            |
Lethality Index for Navy | 11.59         |
Number of Navy officers dead  | 153           |
Number of Navy officers wounded|515           |
Navy officers dead-to-wound ratio| 0.30%         |
Non Armed Forces Dead-to-Wound Ratio given Navy | 56.26         | 17.3
Army              | Not enough data            |
  

```{r}
#ggplot about differences found
```


----

###Discussion about the Lethality Index

There were some significant issues when trying to calculate the ratios:

1. I was unable to calculate the ratio for Army combatants given the lack of 
data available.
  
2. We are unable to see the events in which 100% dead but zero wounded for any category nor can we see the people that were "saved" or ran away and etc...
  
3. It is not easy to calculate the Navy ratio without a comprehensive mapping 
of the location of each confrontation in order to properly count the events in
which the Navy could have participated.
  
4. Missingness of data: (I am not sure if that was intentional or impossible to calculate given the data we have). The lethality index variables in the dataset are mostly blank (see plot below). One potential explanation is that when creating the index (by applying the ratio previously discussed) we have many zero values which prevent the index from being more useful.
  

```{r, warning=FALSE,echo=FALSE}
maping_nas <- aggr(dta, plot = FALSE, col = c('navyblue','red'), prop = TRUE, bars = TRUE, numbers = FALSE, sortVars = TRUE, labels = names(dta), combined = TRUE, only.miss = TRUE, cex.axis = .5, gap = 4, digits = 2, ylab = c("Histogram of missing data","Pattern"))

#sum_missing <- maping_nas$count
sum_missing_name <- maping_nas$missings
colnames(sum_missing_name)[2] <- "Count of missing values"

NAsplot <- sum_missing_name %>%
    filter(`Count of missing values` > 19)
```

Variable  | Count of missing values
--------- | -----------------------
state |	20	
state_abbr |	20	
municipality |	20	
organized_crime_lethality |	19744	
army_lethality |	31463	
navy_lethality |	33787	
federal_police_lethality |	32680	
organized_crime_NewIndex |	19744	
army_NewIndex |	31463	
navy_NewIndex |	33787


----



----

##Conclusion: Improving the index

_Why killed-to-wounded ratios don’t measure lethality_

The lack of information on the total number of participants in each armed engagement is not clear. In other words, the denominator is unknown and "it is therefore very hard to estimate dead/wounded rates per event. It is important to make a distinction between the combat efficiency of armed forces - one index could be # of shots on target / shots fired, and the brutality of the armed forces. It is very difficult to measure the "brutality" of, for example, a federal police office for it is difficult to get accurate data on the number of wounded and dead and attribute the blame to the officer per se.

García-Ponce & Lajous (2016) drives the point home when they explain that "although the use of killed-to-wounded ratios has been suggested as an effective metric to track specific massacres (Coupland and Meddings, 1999), it is a statistic more frequently used to analyze medical attention during armed conflicts than to measure combat efficiency, the severity of civil war, or the use of force." 


Furhtremore, the data does not allows us to understand the context of the confrontation, for example we are unsure how many people were invovolded or the type of weapons used. The calculations measures the consequences for a certain group (e.g. Federal Police) of engaging in a confrontation event. In other words, it measures the ratio of Federal Police officers engaged in a confrontation that died in comparison to other officers that were wounded.
Since violent events can have the cooperation of multiple armed force agencies, to identify if an agency was involved in the violent event we set ( $Federal\space Police \space wounded \space \space != 0$ ). So it becomes hard to say who killed/wounded who. 


Lastly, I believe that to get a comprehensive picture on the brutality issue, there has to be qualitative assessments in order to capture the context of different conflicts. This is also very difficult given the high risk of violent conflicts, so most observerbs belong to either the armed forces or criminal organizations, and civilians are usually not comfortable to accurately report on what they saw as they fear any potential backlash. It is possible that, if the violent event was caught on video, to use the recorded images to make better assessments of the brutality and combat efficiency of the armed forces in Mexicol.

----------------
###Footnotes

^1^  Civilians in this case are all that are not part of the Armed Forces, so it includes the organized crime and government employees in other roles.
  
^2^ Armed forces includes all government agencies used to keep the law and order such as: the Federal Police, the Army, the Navy, among others.

---------------

###References

García-Ponce, O., & Lajous, A. (2016, October 21). Challenges of Measuring an Armed Group’s Lethality: Insights from the Mexican Drug War. Retrieved February 16, 2018, from https://politicalviolenceataglance.org/2016/10/21/challenges-of-measuring-an-armed-groups-lethality-insights-from-the-mexican-drug-war/

